---
title: "WMATA Queue Jump Effectiveness: Quick Visualizations"
output: 
  html_notebook:
    toc: true
    toc_depth: 5
    toc_float: yes
    df_print: paged
    number_sections: false
    theme: simplex
---

#### Summary


```{r message=FALSE, warning=FALSE, include=FALSE}
#Load libraries and basic params in separate script
source("00-master-exploratory.R", local = TRUE)
source("01-readin-exploratory.R", local = TRUE)

#Additional libraries
library(extrafont)
library(scales)
library(patchwork)
#First run only
# extrafont::font_import()
# extrafont::loadfonts(device="win")

#Some parameters 
#Suppress internal viewer for mapview (TRUE during writing, use FALSE on knitting)
viewersupp <- FALSE

#Used for replacing geometry

knitr::opts_chunk$set(
  cache = FALSE,
  highlight = TRUE,
  collapse = TRUE
)

rawnav <-
  rawnav_interim 

s9_stops <- 
  get_route_stops(gtfs_obj,stops,"S9") 

s9_nb_ss_stops <-
  s9_stops %>%
  filter(trip_headsign == "SILVER SPRING STATION", shape_id == "2928")

s9_nb_ss_stops %>%
  st_drop_geometry()
```


#### 1. GPS Accuracy and Heading
GPS pings appear to show movement in and out of curb lanes. For example, at the 16th Street and U St NW intersection in the northbound direction, a one bus run's worth of rawnav GPS records appear to show the bus pulling aside to enter the bus zone. For reference, [Google street view of this location is available here](https://www.google.com/maps/@38.9167036,-77.0365272,3a,75y,49.33h,76.11t/data=!3m6!1e1!3m4!1sYahUtvXy7LQjhGNjKtJkWg!2e0!7i16384!8i8192).

Note that because of map tile layer limitations, you cannot zoom in any further on this map. To see points more clearly, try toggling to a different base map. This stop indeed recorded a door open and close at the point where speeds are lowest in the map below.

```{r}
{rawnav %>%
    filter(id == 48942 &
           (gps_reading_secs_past_midnight >= 26307 |
             gps_reading_secs_past_midnight < 26365)) %>% #this is an S9 NB
  arrange(door_state) %>%
  mapview(.,
          zcol = "mph",
          layer.name = "Speed (mph)",
          map.types = c("Esri.WorldImagery","CartoDB.Positron"))}@map %>%
  setView( #zoom on 16th and U
    lng = -77.0363,
    lat = 38.9169,
    zoom = 18
  )
```


```{r}
{rawnav %>%
    filter(id == 48942 &
           (gps_reading_secs_past_midnight >= 26307 |
             gps_reading_secs_past_midnight < 26365)) %>% #this is an S9 NB
  arrange(door_state) %>%
  mutate(door_state = if_else(door_state == "C","Closed","Open")) %>%
  mapview(.,
          zcol = "door_state",
          layer.name = "Door State",
          map.types = c("Esri.WorldImagery","CartoDB.Positron"))}@map %>%
  setView( #zoom on 16th and U
    lng = -77.0363,
    lat = 38.9169,
    zoom = 18
  )
```


```{r}
{rawnav %>%
    filter(id == 48942 &
           (gps_reading_secs_past_midnight >= 26307 |
             gps_reading_secs_past_midnight < 26365)) %>% #this is an S9 NB
  arrange(door_state) %>%
  mapview(.,
          zcol = "heading_deg",
          layer.name = "Heading",
          map.types = c("Esri.WorldImagery","CartoDB.Positron"))}@map %>%
  setView( #zoom on 16th and U
    lng = -77.0363,
    lat = 38.9169,
    zoom = 18
  )
```


