---
title: "Understanding Start and End Locations in Rawnav data"
output: 
  html_notebook:
    code_folding: hide
    toc: true
    toc_depth: 4
    toc_float: yes
    df_print: paged
    theme: simplex
date: "Updated April 16, 2020"

---

How to handle cases that fail to begin and end near the first and last stop of a route, respectively, could have a significant effect on the amount of data available to analyze. For each run, the rawnav parser identifies the location of the rawnav GPS pings closest to the first and last stops of the route's pattern in GTFS, using these pings to identify the beginning and end of a rawnav route. 

### About the Source Data
This notebook builds upon rawnav data parsed for Route 79 using an in-development version of the rawnav parser.

```{r message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  cache = FALSE,
  highlight = TRUE,
  collapse = TRUE
)

#Load libraries, basic params, and other data in separate script
source("00-master-exploratory.R", local = TRUE)
source("01-readin-exploratory.R", local = TRUE)

path_processed <- file.path(sharepointpath,
                            "Client Shared Folder",
                            "data",
                            "02-processed")

#Prep work on trip summaries
summary79raw <- 
  readxl::read_excel(path = file.path(path_processed,
                                      "Route79_TrimSumv1.xlsx"))

summary79 <-
  summary79raw %>%
  mutate(trip_start_time = lubridate::ymd_hms(paste0("1970-01-01 ",TripStartTime), tz = "EST"),
         trip_end_time = lubridate::ymd_hms(paste0("1970-01-01 ",TripEndTime), tz = "EST"),
         trip_start_time_hr = lubridate::hour(trip_start_time),
         dirno = str_sub(Tag, start = 3),
         dirint = if_else(dirno == "01",
                          1,
                          0),
         dirname = if_else(dirno == "01",
                           "Southbound",
                           "Northbound"),
         time_period = cut(trip_start_time_hr,
                           breaks = c(0,4,6,9,15,19,23,99),
                           labels = c("Late_Night", "AM_Early", "AM_Peak", "Midday", "PM_Peak", "Evening", "Late_Night"),
                           ordered = TRUE,
                           include.lowest = TRUE,
                           right = FALSE),
         refid = glue::glue("{Date} - {TripStartTime} - {dirname} - {FileNm}")) %>%
  make_extra_geom(Long_start_GTFS, Lat_start_GTFS,geom_start,4326) %>%
  make_extra_geom(Long_end_GTFS, Lat_end_GTFS, geom_end, 4326)


#Prep work on GTFS
#Note 79-0 is northbound in GTFS, 79-1 is southbound-- same as rawnav as it happens
route79_nb_shapes <- 
  wmata_shapes %>%
  filter(route_short_name == "79", direction_id == 0) %>%
  summarize()

route79_sb_shapes <- 
  wmata_shapes %>%
  filter(route_short_name == "79", direction_id == 1) %>%
  summarize()

 route79_stops <- 
    gtfs_obj$routes %>%
    filter(route_short_name == "79") %>%
    left_join(gtfs_obj$trips, by = "route_id") %>%
    left_join(gtfs_obj$stop_times, by = "trip_id") %>%
    group_by(route_short_name,direction_id,trip_headsign,shape_id,stop_id) %>%
    summarize(trips = length(unique(trip_id)),
              #not a great assumption, but this is a quickie
              stop_sequence = max(stop_sequence)) %>%
    left_join(stops, by = "stop_id") %>%
    arrange(stop_sequence, .by_group = TRUE) %>%
    ungroup() %>%
    st_sf() 

#Prep work on distance to a stop
calc_dist <- function(geom_col,
                      dir = 0,
                      stopsfile,
                      where = "start"){
  #this is intentionally a bit thrown together, quickly recalculating to 
  #get notebook out the door
  stopifnot(dir %in% c(0,1))
  stopifnot("sf" %in% class(stopsfile))
  stopifnot(where %in% c("start","end"))
  if (where == "start"){
    check_stop <- 
      stopsfile %>%
      filter(direction_id == dir) %>%
      #intentionally a little sloppy here since patterns on 79 are simple
      filter(stop_sequence == min(stop_sequence)) %>%
      filter(row_number() ==1)
  } else {
    check_stop <- 
      stopsfile %>%
      filter(direction_id == dir) %>%
      filter(stop_sequence == max(stop_sequence)) %>%
      filter(row_number() ==1)
  }

  stopifnot(nrow(check_stop)==1)
  
  st_distance(st_set_crs(st_sfc(geom_col),4326),
              check_stop) %>%
    units::set_units(.,value = "ft")
  
}
 
 
summary79_2 <-
  summary79 %>%
  mutate(dist_to_first = pmap(list(geom_start,dirint),
                              calc_dist,
                              route79_stops,
                              where = "start"),
         dist_to_last = pmap(list(geom_end,dirint),
                             calc_dist,
                             route79_stops,
                             where = "end"))

summary79_3 <-
  summary79_2 %>%
  mutate(dist_to_first = unlist(dist_to_first),
         dist_to_last = unlist(dist_to_last))
 
```

The data here includes 411 bus runs. This is a small subset of the data available for Route 79 for the month of October 2019.

The runs themselves span the entire month, with most files including several runs of Route 79.

```{r}
summary79_3 %>%
  count(Date,FileNm) %>%
  arrange(Date,FileNm)
```

The runs are present across multiple directions and time periods. 

```{r}
summary79_3 %>%
  count(time_period, dirname) %>%
  spread(time_period,n)
  
```

Not all of the runs appear to run the full length of the alignment. A handful of runs have zero distance, and others are noticeably short.
```{r}
summary79_3 %>%
  ggplot(aes(x = Trip_Dist_Mi_GTFS)) +
  ggtitle("Distribution of Running Distances for Route 79 Runs") +
  geom_histogram() +
  FITP_Theme
```
In the examination of Route 79 below, runs of less than 3 miles (slightly less than half the length of the alignment) are excluded from the analysis. This results in `r summary79 %>% filter(Trip_Dist_Mi_GTFS >= 3) %>% nrow()` remaining records out of `r nrow(summary79)`.

### Trip Geometry

#### Start Points
Notably, not all of the trips begin where one might expect. First, northbound rawnav origin points and the underlying route. The GTFS first and last stops in the northbound direction are shown with markers. 

**79 Northbound Rawnav Trip Start Points**
```{r message=FALSE, warning=FALSE}
summary79_3 %>%
  st_set_geometry("geom_start") %>%
  filter(dirname == "Northbound") %>%
  filter(Trip_Dist_Mi_GTFS >= 3) %>%
  mapview(zcol = "Dist_from_GTFS1stStop_start_ft", 
          layer.name = "Dist. From First Stop",
          homebutton = TRUE) +
  mapview(route79_nb_shapes, legend = FALSE) +
  mapview({
    route79_stops %>%
      filter(direction_id == 0) %>%
      filter(stop_sequence == min(stop_sequence))
  },
  color = "red",
  col.regions = "red",
  alpha.regions = 1,
  legend = TRUE,
  layer.name = "Route 79 NB First GTFS Stop")
```
Notably, in the southbound direction, there are a number of runs that do not 'begin' until near the end of the 79 SB alignment. These runs are filtered out in the map below based on the minimum running distance criteria of 3 miles.
**79 Southbound Rawnav Trip Start Points**
```{r}
summary79_3 %>%
  st_set_geometry("geom_start") %>%
  filter(dirname == "Southbound") %>%
  filter(Trip_Dist_Mi_GTFS >= 3) %>%
  mapview(zcol = "Dist_from_GTFS1stStop_start_ft", 
          layer.name = "Dist. From First Stop",
          homebutton = TRUE) +
  mapview(route79_sb_shapes, legend = FALSE) +
  mapview({
    route79_stops %>%
      filter(direction_id == 1) %>%
      filter(stop_sequence == min(stop_sequence))
  },
  color = "red",
  col.regions = "red",
  alpha.regions = 1,
  legend = TRUE,
  layer.name = "Route 79 SB First GTFS Stop")
```
#### End Points
In the Northbound direction, it appears some routes never quite start, or start late.

**79 Northbound Rawnav Trip End Points**
```{r}
summary79_3 %>%
  st_set_geometry("geom_end") %>%
  filter(dirname == "Northbound") %>%
  filter(Trip_Dist_Mi_GTFS >= 3) %>%
  mapview(zcol = "Dist_from_GTFSlastStop_end_ft", 
          layer.name = "Dist. From Last Stop",
          homebutton = TRUE) +
  mapview(route79_nb_shapes, legend = FALSE) +
  mapview({
    route79_stops %>%
      filter(direction_id == 0) %>%
      filter(stop_sequence == max(stop_sequence))
  },
  color = "red",
  col.regions = "red",
  alpha.regions = 1,
  legend = TRUE,
  layer.name = "Route 79 NB Last GTFS Stop")
```

Southbound, it appears other runs never fully depart, and some may not close out appropriately. 

**79 Southbound Rawnav Trip End Points**
```{r}
summary79_3 %>%
  st_set_geometry("geom_end") %>%
  filter(dirname == "Southbound") %>%
  filter(Trip_Dist_Mi_GTFS >= 3) %>%
  mapview(zcol = "Dist_from_GTFSlastStop_end_ft", 
          layer.name = "Dist. From Last Stop",
          homebutton = TRUE) +
  mapview(route79_nb_shapes, legend = FALSE) +
  mapview({
    route79_stops %>%
      filter(direction_id == 1) %>%
      filter(stop_sequence == max(stop_sequence))
  },
  color = "red",
  col.regions = "red",
  alpha.regions = 1,
  legend = TRUE,
  layer.name = "Route 79 SB Last GTFS Stop")
```
#### Cases with Issues
```{r message=FALSE, warning=FALSE, include=FALSE}
usefil <- 500
```

What of the cases that fall further than `r usefil` feet away from a start or end point? Because rawnav pings are not loaded in this dataset, only the straightlines from the start to the end are shown. Many of these trips are making the journey. Here we look just at the Northbound trips, which had fewer close out issues previously. 

**79 Northbound Trips With First/Last Stop Issues** 
```{r}
summary79_4 <-
  summary79_3 %>%
  select(Long_start_GTFS,Lat_start_GTFS,Long_end_GTFS,Lat_end_GTFS) %>%
  stplanr::od_coords2line(remove_duplicates = FALSE) %>%
  select(geometry) %>%
  {bind_cols(summary79_3,.)} %>%
  st_set_geometry("geometry") %>%
  mutate(length = st_length(geometry)) %>%
  mutate(refid = as.character(refid)) 


summary79_4 %>%
  filter(dirname == "Northbound") %>%
  filter(dist_to_first > usefil | dist_to_last > usefil) %>%
  mapview(.,
          zcol = "refid", 
          layer.name = "Run Info",
          legend = FALSE,
          alpha = .4,
          highlight = highlightOptions(bringToFront = TRUE,
                                       weight = 5,
                                       fillOpacity= 1,
                                       color="red"),
          homebutton = TRUE) 
```

### Run Times
If we filter to runs with rawnav pings that fall within a set distance from the first *and* last stop, how many records do we get?
```{r}
summary79_fil <-
  summary79_3 %>%
  filter(dist_to_first <= usefil & dist_to_last <= usefil) %>%
  filter(Trip_Dist_Mi_GTFS >= 3) %>%
  mutate(runtime = Trip_Dur_Sec_GTFS / 60) 

```
Using `r usefil` feet, we keep `r nrow(summary79_fil)` records, or `r scales::percent(nrow(summary79_fil) / nrow(summary79))` of the sample we started with.

By time period, we see the following counts:
```{r}
summary79_fil %>%
  count(time_period,dirname) %>%
  spread(time_period,n)
```

What's the distribution of runtimes? Below are averages by period, in minutes. These are calculated based on time intervals between the pings closest to the first and last stops.
```{r}
summary79_fil %>%
  group_by(time_period,dirname) %>%
  summarize(runtime = round(mean(runtime, na.rm = TRUE),2)) %>%
  spread(time_period,runtime)
```

There are some outliers for runtime.
```{r}
summary79_fil %>%
  ggplot(aes(x = runtime)) +
  ggtitle("Distribution of Runtimes for Filtered Route 79 Trips") +
  geom_histogram() +
  FITP_Theme
```


