---
title: "WMATA Queue Jump Effectiveness: Decomposition Method Comparison"
output: 
  html_notebook:
    toc: true
    toc_depth: 2
    toc_float: yes
    df_print: paged
    number_sections: false
    theme: simplex
---



```{r message=FALSE, warning=FALSE, echo = FALSE}
source("./decomposition-prep.R")
source("./decomposition-sep-accel-decel-calcs.R")
source("./decomposition-compare-df.R")

```

We'll look at two methods:
* Traveltime Decomposition (what was presented at methodology workshop with 14 second placeholder for t_stop2)
* Alternative Decomposition (method that breaks down accel/decel phases and in-stop-area/out-of-stop-area separately, referred to as the A/D method)

In both cases, let's filter out cases where segment runtimes are over 500 seconds, then pick cases where the two have segment second totals that are the same (within 2 seconds, given rounding that occurs in some places that can be removed later). A few issues would still need to be addressed if the alternative decomposition is used that haven't been implemented for now -- namely, segments with multiple stops are more complicated and a chunk of R code would need to be updated/replaced in Python to produce accurate values for some runs (especially on Irving at 15th/16th). The runs that are shown have accurate values that can be compared across methods.

Behind the scenes we'll load data and combine the results of the tables together.
```{r, echo = FALSE}
tt_decomp_fil <-
  tt_decomp %>%
  filter(secs_seg_total < 500) %>%
  unite(busrun, filename, index_run_start) %>%
  semi_join(compare_df, by = c("busrun", "seg_name_id"))
  
ad_decomp_fil <-
  stop_pass_decomp_4 %>%
  semi_join(tt_decomp_fil, by = c("busrun", "seg_name_id")) %>%
  mutate_if(is.numeric, ~ replace_na(., 0)) %>%
  group_by(seg_name_id) %>%
  mutate(
    avg_approach_delay = mean(ts_approach_min + tr_approach_delay, na.rm = TRUE),
    avg_accel_delay = mean(ts_accel_min + tr_accel_delay, na.rm = TRUE),
  ) %>%
  mutate(
    t_qj_related =
      subsegment_delay_sec_after_stop_area +
      tr_accel_delay,
    t_ad_decomp_secs_total =
      tot_secs_check +
      subsegment_min_sec_after_stop_area +
      subsegment_min_sec_before_stop_area +
      subsegment_delay_sec_after_stop_area +
      subsegment_delay_sec_before_stop_area,
    t_seg_ad_reclac = tot_secs_check + 
              subsegment_min_sec_after_stop_area +
              subsegment_min_sec_before_stop_area +
              subsegment_min_sec_between_stop_area +
              subsegment_delay_sec_between_stop_area +
              subsegment_delay_sec_after_stop_area +
              subsegment_delay_sec_before_stop_area
  )

```

```{r, echo = FALSE}
compare_fil <-
  tt_decomp_fil %>%
  select(
    busrun,
    seg_name_id,
    route,
    wday,
    start_date_time,
    t_stop,
    t_stop1,
    t_stop2,
    t_ff,
    t_traffic,
    secs_seg_total
  ) %>%
  left_join(
    {select(ad_decomp_fil,
           busrun,
           seg_name_id,
           ts_approach_min:subsegment_delay_sec_before_stop_area,
           t_qj_related,
           t_seg_ad_reclac) %>%
     select(-ts_nostop_min,
            -tr_nostop_delay,
            -checksout)
    },
    by = c("busrun","seg_name_id"),
    suffix = c("_tt","_ad")
  ) %>%
  mutate(total_check = secs_seg_total - t_seg_ad_reclac) %>%
  filter(abs(total_check) < 2)

```

First, let's check that t_stop1 is same in both cases
```{r, echo = FALSE}
testthat::test_that("tstop1 matches",{
  test_bad_t_stop1 <-
    compare_fil %>%
    mutate(t_stop1_check = t_stop1 == tr_stop1) %>%
    filter(!t_stop1_check)
  
  testthat::expect_equal(nrow(test_bad_t_stop1),0)
})

```

WE'll compare the two results on two values

* t_traffic: From the proposed decomposition.
* t_qj_related: From the alternative decomposition, a shorthand for acceleration delay in the stop area and delay in the remainder of the segment after the stop. The presence of a QJ might plausibly be related to this time.

It looks like for a range of t_qj_related values, there's a wide range of traffic related values. For instance, at 100 seconds of t_qj_delay, we can get as much as 300 seconds of t_traffic. Before we dive in, we'll look at a few other instances. Notably, both have some below-zero values. It also appears that differences may be smaller over longer segments as the effect of areas outside the segment dwarf the differences near the stop area.
```{r}
gg <- compare_fil %>%
  plotly::highlight_key(., ~busrun) %>%
  ggplot(., aes(x = t_qj_related,
                y = t_traffic)) +
  geom_point() 

gg %>% ggplotly()
# 
# gg
```
In general though, they're pretty similar. R2 is about 85%
```{r}
cor(compare_fil$t_qj_related, compare_fil$t_traffic)
```

Looking by segment, nothing immediately jumps out. Note that a few segments have been dropped because there are some remaining issues with the decomposition in these segments where total segment runtime doesn't exactly match because of updates that must be made to the alternative decomposition.

```{r}
gg2 <- compare_fil %>%
  plotly::highlight_key(., ~busrun) %>%
  ggplot(., aes(x = t_qj_related,
                y = t_traffic, 
                color = seg_name_id)) +
  geom_point(
         alpha = .2,
         shape = 21,
         fill = NA
  ) 

gg2
```

What if we filter to weekday PM peak? We still see a bit of a distribution

```{r}
compare_fil_pmpeak <-
  compare_fil %>%
  filter(!(wday %in% c("Saturday","Sunday")), 
         lubridate::hour(start_date_time) >= 16, 
         lubridate::hour(start_date_time) < 17)

gg3 <- compare_fil_pmpeak %>%
  plotly::highlight_key(., ~busrun) %>%
  ggplot(., aes(x = t_qj_related,
                y = t_traffic, 
                color = seg_name_id)) +
  geom_point(
         alpha = .4,
         shape = 21,
         fill = NA
  ) 

gg3
```

Let's look at a few particular cases before we dive into an explanation. Here's one that has about 97 secs related to queue jump under the A/D version, but 180 seconds of t_traffic. A big difference seems to be the 58 seconds of delay in the stop zone before reaching the stop ('tr_approach_delay'). THat gets us most of the way there. 

```{r}
case_run = "rawnav06455191013.txt_8278"
case_seg = "sixteenth_u_long"

compare_fil %>%
  filter(busrun == "rawnav06455191013.txt_8278") %>% 
  base::t()

```

```{r}
x <- doublecheck(case_run, case_seg)
x
```

What about one with a small difference? 

There's 126 seconds of QJ related delay here via the A/D method, versus about 135 seconds of t_traffic delay. Notably, there's far less delay before the stop. In this way, the two line up well.

```{r}
case_run = "rawnav06467191019.txt_10039"
case_seg = "sixteenth_u_long"

compare_fil %>%
  filter(busrun == case_run) %>% 
  base::t()

```

```{r}
x <- doublecheck(case_run, case_seg)
x
```

What about a case with very little delay addressed by the queue jump and lots of traffic delay? 

Here, there's 13 seconds of delay associated with acceleration or downstream delay, vs about 187 seconds of t_traffic delay.

```{r}
case_run = "rawnav06027191014.txt_2752"
case_seg = "sixteenth_u_shrt"

compare_fil %>%
  filter(busrun == case_run) %>% 
  base::t()

```

Here, the big difference is associated with signal delay, or some bus behavior that looks like that. That value is reflected in t_traffic but not in the combination of acceleration delay and downstream delay referred to as t_qj_related.
```{r}
x <- doublecheck(case_run, case_seg)
x
```


```{r}
compare_fil %>%
  write_csv("compare_methods_by_run.csv")
```

This case appears to illustrate that some acceleration delay is in fact signal delay, given the long time before acceleration 'really' begins.

```{r}
case_run = "rawnav02801191018.txt_7426"
case_seg = "sixteenth_u_shrt"

compare_fil %>%
  filter(busrun == case_run) %>% 
  base::t()


```

```{r}
x <- doublecheck(case_run, case_seg)
x
```

