---
title: "WMATA Queue Jump Effectiveness: Decomposition Using Separate Accel/Decel, Min+Delay Method"
output: 
  html_notebook:
    toc: true
    toc_depth: 2
    toc_float: yes
    df_print: paged
    number_sections: false
    theme: simplex
---

This notebook illustrates a possible decomposition of travel time focused on separate accel/decel speeds to illustrate the minimum and maximum of each.

We load the same data as the decomposition early-look notebook.
```{r message=FALSE, warning=FALSE}
source("./decomposition-prep.R")

```

## Decomposition

We continue with a decomposition based on the inputs thus far. For now, we dump this code in, will break out components and discuss later.


```{r}
calc_phase_delay <- function(rawnav_stop_area,
                             phase = "t_accel_phase",
                             name = "accel"){
  
  #TODO: if in the 'no stop' case, maybe run 95th percentile for all trips
  # and then difference this.
  
  speed_1 <-
    rawnav_stop_area %>%
    filter(stop_area_phase %in% phase) %>%
    group_by(busrun, seg_name_id) %>%
    summarize("tot_{name}_secs" := sum(secs_marg, na.rm = TRUE),
              "tot_{name}_odom_ft" := sum(odom_ft_marg, na.rm = TRUE)) %>%
    ungroup() %>%
    # to be safe, we set any fps values of NaN or Inf to NA -- probably a case where 
    # the stop zone is defined too tightly and we get weird cases with inf speed
    mutate("{name}_fps" := 
             .data[[glue::glue("tot_{name}_odom_ft")]] 
           / na_if(.data[[glue("tot_{name}_secs")]],0)) 
  
  speed_1_distr <-
    speed_1 %>%
    group_by(seg_name_id) %>%
    summarize(distr = list(quantile(.data[[glue::glue("{name}_fps")]], 
                                    probs = c(.02,.05,.1,.25,.5,.75,.9,.95,.98),
                                    na.rm = TRUE))) 
  
  low_speed_to_join <- 
    speed_1_distr %>%
    unnest_longer(col = distr,
                  values_to = glue::glue("{name}_fps"),
                  indices_to  = "ntile") %>%
    filter(ntile == "95%") %>%
    select(seg_name_id, 
           "low_{name}_fps" := .data[[glue::glue("{name}_fps")]])
  
  speed_2 <-
    speed_1 %>%
    left_join(low_speed_to_join, by = "seg_name_id") %>%
    mutate("min_{name}_secs" := .data[[glue::glue("tot_{name}_odom_ft")]] / .data[[glue::glue("low_{name}_fps")]],
           "addl_{name}_secs" := .data[[glue::glue("tot_{name}_secs")]] - .data[[glue::glue("min_{name}_secs")]]) %>%
    select(busrun, seg_name_id,glue::glue("min_{name}_secs"), glue::glue("addl_{name}_secs"))
  
}

# No stop phase -------
no_stop_time <- calc_phase_delay(rawnav_stop_area,
                               "t_nostopnopax",
                               "nostopnopax")

# Decel Phase -----
decel_time <- calc_phase_delay(rawnav_stop_area,
                               "t_decel_phase",
                               "decel")

# Accel Phase -----
accel_time <- calc_phase_delay(rawnav_stop_area,
                                 "t_accel_phase",
                                 "accel")

# bring it all together ----
stop_pass_decomp_1 <- 
  rawnav_stop_area %>%
  group_by(busrun, seg_name_id, stop_area_phase) %>%
  summarize(tot_secs = sum(secs_marg, na.rm = TRUE)) %>%
  ungroup() %>%
  pivot_wider(names_from = "stop_area_phase", values_from = "tot_secs") %>%
  rowwise() %>%
  mutate(tot_secs =
           sum(
           t_decel_phase, 
           t_l_initial,
           t_stop1,
           t_stop,
           t_l_addl,
           t_accel_phase,
           t_nostopnopax,
           na.rm = TRUE
           )) %>%
  ungroup() %>%
  select(busrun,
         seg_name_id,
         t_decel_phase,
         t_l_initial,
         t_stop1,
         t_stop,
         t_l_addl,
         t_accel_phase,
         t_nostopnopax,
         tot_secs)

stop_pass_decomp_2 <-
  stop_pass_decomp_1 %>%
  left_join(no_stop_time, by = c("busrun","seg_name_id")) %>%
  left_join(decel_time,   by = c("busrun","seg_name_id")) %>%
  left_join(accel_time,   by = c("busrun","seg_name_id")) 

stop_pass_decomp_3 <-
  stop_pass_decomp_2 %>%
  mutate(ts_approach_min = min_decel_secs,
         tr_approach_delay = addl_decel_secs,
         tr_init_wait = t_l_initial,
         tr_stop = t_stop,
         tr_stop1 = t_stop1,
         tr_signal_wait = t_l_addl,
         ts_accel_min = min_accel_secs,
         tr_accel_delay = addl_accel_secs,
         ts_nostop_min = min_nostopnopax_secs,
         tr_nostop_delay = addl_nostopnopax_secs
         ) %>%
  rowwise() %>%
  mutate(tot_secs_check = 
           sum(
             ts_approach_min, 
             tr_approach_delay,
             tr_init_wait,
             tr_stop1,
             tr_stop,
             tr_signal_wait,
             ts_accel_min,
             tr_accel_delay,
             ts_nostop_min,
             tr_nostop_delay,
             na.rm = TRUE)
  ) %>%
  ungroup() %>%
  mutate(
    tot_secs_check = round(tot_secs_check),
    checksout = tot_secs == tot_secs_check
    )
 
rawnav_nonstop_area_wide <-
  rawnav_nonstop_area %>%
  pivot_wider(id_cols = busrun:seg_name_id,
              names_from = segment_part,
              values_from = subsegment_min_sec:subsegment_delay_sec)

stop_pass_decomp_4 <- 
  stop_pass_decomp_3 %>%
  left_join(rawnav_nonstop_area_wide, by = c("busrun","seg_name_id"))


# Test that the 
testthat::test_that("check sum matches for everything",{
  # TODO: go fix
  doesnt_checkout <-
    stop_pass_decomp_4 %>%
    filter(!checksout)
  
  testthat::expect_true(nrow(doesnt_checkout) == 0)
})

```


# Create Charts
```{r}
decomp_stack <- 
  function(decomp){
  # set cols
    ltblue = "#A6CEE3"
    dkblue = "#1F78B4"
    ltgrn = "#B2DF8A"
    dkgrn = "#33A02C"
    ltred = "#FB9A99"
    dkred = "#E31A1C"
    ltorg = "#FDBF6F"
    dkorg = "#FF7F00"  
    # browser()
  # data check
  chartdata <-
    decomp %>%
    select(
      busrun,
      seg_name_id,
      ts_approach_min,
      tr_approach_delay,
      tr_init_wait,
      tr_stop1,
      tr_stop,
      tr_signal_wait,
      ts_accel_min,
      tr_accel_delay,
      ts_nostop_min,
      tr_nostop_delay
    ) %>%
    pivot_longer(
      names_to = "decomp_state",
      values_to = "secs",
      cols = ts_approach_min:tr_nostop_delay
    ) %>%
    mutate(secs = round(secs,0),
           decomp_state = 
             factor(decomp_state,
                    levels = c("ts_approach_min",
                               "tr_approach_delay",
                               "tr_init_wait",
                               "tr_stop1",
                               "tr_stop",
                               "tr_signal_wait",
                               "ts_accel_min",
                               "tr_accel_delay",
                               "ts_nostop_min",
                               "tr_nostop_delay")
                    ),
           source = if_else(decomp_state %in% c("ts_approach_min",
                                                "ts_accel_min",
                                                "ts_nostop_min"),
                            "Segment-level",
                            "Run-level")
    ) 

  # make cahrt
    ggplot(chartdata,
           aes(x = busrun,
               y = secs, 
               group = busrun, 
               fill = decomp_state,
               color = source,
               label = secs)) + 
      geom_col(size = 1) + 
      geom_label(position = position_stack(vjust = .5)) +
      scale_fill_manual(
        values = c(
        "ts_approach_min" = dkred,
        "tr_approach_delay" = ltred,
        "tr_init_wait" = ltblue, 
        "tr_stop1" = dkblue,
        "tr_stop" = dkblue,
        "tr_signal_wait" = ltorg,
        "ts_accel_min" = dkgrn,
        "tr_accel_delay" = ltgrn,
        "ts_nostop_min" = dkred,
        "tr_nostop_delay" = ltred
        )
      ) + 
      scale_color_manual(
        values = c("Segment-level" = "black",
                   "Run-level" = "#6a3d9a" )
      ) +
      guides(color = guide_legend(reverse = TRUE)) +
      labs(y = "Seconds")
    
  }

doublecheck <- function(case, seg){

  dc <- 
    stop_pass_decomp_4 %>%
    filter(busrun == case, seg_name_id == seg) %>%
    decomp_stack()
  
  ts <- rawnav_stop_area %>%
    filter(busrun == case, seg_name_id == seg) %>%
    make_basic_timespace()
  
  sd <- rawnav_stop_area %>%
    filter(busrun == case, seg_name_id == seg) %>%
    make_basic_speeddist(alpha = .5, 
                         size = 1.5, 
                         lt = "solid",
                         legend = "none")

  nonstoparea_subseg_table <- 
    rawnav_nonstop_area %>%
    filter(busrun == case, seg_name_id == seg) %>%
    mutate(segment_part = str_replace_all(segment_part,
                                          c("after_stop_area" = "After Stop Area",
                                            "before_stop_area" = "Before Stop Area"))) %>%
    select(-busrun, -seg_name_id) %>%
    gridExtra::tableGrob()
  
  bigplot <-
    (((ts / sd) | dc)) +
    plot_annotation(glue::glue("{seg}: Results for run {case}"))
  
  bigplot / nonstoparea_subseg_table
}

# DEBUG SPEED STUFF
# case <- "rawnav03235191030.txt_26551"
# case <- "rawnav03259191029.txt_28005"
# case <- "rawnav03232191024.txt_9587"
# 

stop_pass_decomp_4 %>%
  filter(busrun == "rawnav03232191024.txt_9587") %>%
  filter(seg_name_id == first(seg_name_id)) %>%
  decomp_stack()


```

```{r}
checkme <- 
  stop_pass_decomp_4 %>%
  filter(busrun == "rawnav03215191019.txt_11241") %>%
  filter(seg_name_id == first(seg_name_id)) 

doublecheck("rawnav02833191007.txt_1494", "irving_fifteenth_sixteenth")

```


# Export Charts
```{r}
frame <-
  rawnav_stop_area %>%
  distinct(busrun, seg_name_id) %>%
  sample_n(size = 500) %>%
  mutate(
    busrunshow = str_replace(busrun, '\\.txt', 'txt'),
    filename = glue::glue("{seg_name_id}_{busrunshow}.png"),
    plot = pmap(list(busrun, seg_name_id),
                doublecheck)
  )

height = 8
asp = 1.2

frame %>%
  select(filename,
         plot) %>%
  pwalk(.,
        ggsave,
        path = file.path(path_project,'data','01-interim','images'),
        width = height * asp,
        height = height,
        units = "in",
        scale = .9,
        dpi = 300)
```

```{r}
stop_pass_decomp_4 %>%
  select(
      busrun,
      seg_name_id,
      ts_approach_min,
      tr_approach_delay,
      tr_init_wait,
      tr_stop1,
      tr_stop,
      tr_signal_wait,
      ts_accel_min,
      tr_accel_delay,
      ts_nostop_min,
      tr_nostop_delay,
      subsegment_min_sec_before_stop_area,
      subsegment_delay_sec_before_stop_area,
      subsegment_min_sec_after_stop_area,
      subsegment_delay_sec_after_stop_area
    ) %>%
  write_csv(.,file.path(path_data,'acceldecel_decomp_byrun.csv'),na = "")
```

