---
title: "WMATA Queue Jump Effectiveness: Free Flow Speed Checks"
output: 
  html_notebook:
    toc: true
    toc_depth: 2
    toc_float: yes
    df_print: paged
    number_sections: false
    theme: simplex
---

```{r include=FALSE}
source("00-master-exploratory.R", local = TRUE)
path_gtfs <-  file.path(datadir,
                     "wmata-2019-05-18 dl20200205gtfs.zip")
source("02-gtfsprep-exploratory.R", local = TRUE)
```

# Read-in

We'll use a set of Route 79 data for the month of October 2019 to examine free flow speeds over two segments of Georgia Ave: one near Piney Branch and one near Columbia Rd.
```{r message=FALSE, include=FALSE}
rawnav79_raw <-
  read_csv(
    file = file.path(
      sharepointpath,
      "Client Shared Folder",
      "data",
      "01-interim",
      "Route79_20200519.csv"
    ),
    col_types = cols(
      X1 = col_double(),
      IndexLoc = col_double(),
      Lat = col_double(),
      Long = col_double(),
      Heading = col_double(),
      DoorState = col_character(),
      VehState = col_character(),
      OdomtFt = col_double(),
      SecPastSt = col_double(),
      SatCnt = col_double(),
      StopWindow = col_character(),
      Blank = col_double(),
      LatRaw = col_double(),
      LongRaw = col_double(),
      RowBeforeAPC = col_double(),
      route_pattern = col_double(),
      pattern = col_character(),
      IndexTripStartInCleanData = col_double(),
      IndexTripEndInCleanData = col_double(),
      filename = col_character(),
      StartDateTime = col_datetime(format = ""),
      `__index_level_0__` = col_double(),
      wday = col_character(),
      route = col_double()
    ))

#we'll clean this up a bit
rawnav79 <-
  rawnav79_raw %>%
  lazy_dt() %>%
  filter(route_pattern == "7901") %>%
  mutate(tag_starthour = lubridate::hour(StartDateTime)) %>%
  # note, i'm notsure this is a perfect way to group, seeing some potential
  #issues
  group_by(filename,IndexTripStartInCleanData) %>%
  mutate(wday = factor(wday,
                       levels = c("Monday","Tuesday","Wednesday","Thursday","Friday","Saturday","Sunday"),
                       ordered = TRUE),
           hit_stop = str_extract(StopWindow,"E\\d*"),
         stop_segment = hit_stop) %>%
  collect() %>%
  fill(stop_segment, .direction = "down")

rm(rawnav79_raw)

# Our universe of files from another export we did; used as a check on our 
# import here
rawnav_inventory_filtered_79_raw <-
  read_csv(
    file = file.path(
      sharepointpath,
      "Client Shared Folder",
      "data",
      "01-interim",
      "rawnav_inventory_filtered_79.csv"
    ),
    col_types = 
      cols(
        X1 = col_double(),
        fullpath = col_character(),
        filename = col_character(),
        file_busid = col_double(),
        file_id = col_character(),
        taglist = col_character(),
        line_num = col_double(),
        route_pattern = col_character(),
        tag_busid = col_double(),
        tag_date = col_datetime(format = ""),
        tag_time = col_time(format = ""),
        Unk1 = col_double(),
        CanBeMiFt = col_character(),
        route = col_character(),
        pattern = col_character(),
        tag_datetime = col_datetime(format = ""),
        tag_starthour = col_double(),
        wday = col_character()
      ))

rawnav_inventory_filtered_79 <-
  rawnav_inventory_filtered_79_raw %>%
  select(filename,route_pattern, line_num,tag_busid,tag_date,tag_time,tag_datetime,tag_starthour,wday)%>%
  filter(route_pattern == "7901") %>%
  mutate(wday = factor(wday,
                       levels = c("Monday","Tuesday","Wednesday","Thursday","Friday","Saturday","Sunday"),
                       ordered = TRUE))

# Intersections
# From: https://opendata.dc.gov/datasets/intersection-points/data
intersections_raw <- 
  read_sf(dsn = file.path(
      sharepointpath,
      "Client Shared Folder",
      "data",
      "00-raw",
      "Intersection_Points-shp",
      "e3661aa5-95cd-4dfb-b189-f9a3bf607bc82020413-1-7zs0bm.h1bpb.shp"
    )) %>%
  st_transform(DCCRS)

intersections <-
  intersections_raw %>%
  mutate(study = OBJECTID %in% c(3234, # "COLUMBIA ROAD NW AND GEORGIA AVENUE NW",
                                 13324, # "PINEY BRANCH ROAD NW AND GEORGIA AVENUE NW"
                                 10991)) %>%#"GEORGIA AVENUE NW AND IRVING STREET NW"
  clean_names()                                 

# 79 -1 is southbound, direction of our QJs

# Route shapes itesm
# get 79 SB shapes -- there are a few but they're basically identical, 
# so we'll just grab the first one. 

wmata_shapes <- 
  read_sf(dsn = file.path(
      sharepointpath,
      "Client Shared Folder",
      "data",
      "00-raw",
      "Metro_Bus_Lines-shp",
      "5606f7ae-68fa-42f5-95c9-9397892d1d552020412-1-19boies.o2k1.shp"
    ))%>%
  st_transform(DCCRS)


shape_7901 <-
  wmata_shapes %>%
  filter(RT_D == "79_S") %>%
  clean_names() 

stops_7901 <-
  get_route_stops(
    gtfs_obj,
    stops,
    "79",
    1
  )

```

This data contains `r scales::comma(nrow(rawnav79))` rows. A sample is shown below. Other metadata about these runs is not included in the table below.

```{r}
head(rawnav79)
```

```{r}
files_pulled <-
  rawnav79 %>%
  lazy_dt() %>%
  distinct(filename) %>%
  collect() 
```

`r nrow(files_pulled)` distinct rawnav zip files are included in this dataset. Each rawnav zip file included will contain some 79 NB, 79SB, pull-in, pull-out, and records for some other routes that are on the same block as a 79 trip. We have filtered to 79 southbound runs, but have not yet filtered this universe of runs to exclude ones that have AVL quality issues.

```{r}
runs_pulled <-
  rawnav79 %>%
  lazy_dt() %>%
  filter(route_pattern == "7901") %>%
  distinct(filename,IndexTripStartInCleanData) %>%
  collect()

```
Around `r scales::comma(nrow(runs_pulled))` unique runs of the 79 southbound are included in this datset out of `r total <- nrow(distinct(rawnav_inventory_filtered_79,filename,route_pattern, tag_date,tag_time, tag_busid)); scales::comma(total)` runs that are in the October 2019 data: close to `r scales::percent(nrow(runs_pulled) / total)`. We'll investigate what happened to that missing run at some point.

The distribution of trips by time of day and day of week are shown below. Start hour is shown on the x axis, day of week on the y-axis, and the number of bus runs in the dataset in each cell.

```{r message=FALSE, warning=FALSE}
df <- 
  rawnav79 %>%
  lazy_dt() %>%
  group_by(
    filename,
    IndexTripStartInCleanData,
    tag_starthour, 
    wday) %>%
  summarize() %>%
  ungroup() %>%
  count(tag_starthour, wday, .drop = TRUE) %>%
  collect() %>%
  #factors aren't playing well with dtplyr, doing this again
  mutate(wday = factor(wday,
                       levels = c("Monday","Tuesday","Wednesday","Thursday","Friday","Saturday","Sunday"),
                       ordered = TRUE))

make_heat <- function(dfin, val, text){
  ggplot(dfin, aes(wday, tag_starthour)) + 
    geom_tile(aes(fill = {{val}})) +
    scale_fill_continuous(type = "viridis") +
    geom_text(aes(label = {{text}}),
              color = "grey80") +
    scale_y_reverse('Trip Start Hour',breaks = 4:21) +
    labs(x = "Day of Week") +
    guides(fill = guide_legend(title="Count of Trips",
                               reverse = TRUE)) + 
    FITP_Theme
}

make_heat(df,n,n)

```


By comparison, the distribution of all 79 southbound runs by day of week and hour for October 2019 is shown below.
```{r message=FALSE, warning=FALSE}
df2 <- 
rawnav_inventory_filtered_79 %>%
 count(tag_starthour, wday)

make_heat(df2,n,n)

```

The percentage of trips captured in each day and hour are shown below.

```{r message=FALSE, warning=FALSE}
df3 <- 
  df2 %>%
  left_join(df, 
            by = c("tag_starthour", "wday"),
            suffix = c(".total",".sample")) %>%
  mutate(percent = n.sample / n.total,
         percent_fmt = scales::percent(percent, accuracy = 1)) %>%
  drop_na(percent)

make_heat(df3,percent,percent_fmt) +
      guides(fill = guide_legend(title="Percent of Trips"))
  
```

# Defining boundaries for our segment analyses

First, where are our DC intersections around the 79 southbound?

```{r}
shape_7901_buff <- st_buffer(shape_7901, dist = units::set_units(100,"ft"))

intersections_7901 <-
  intersections %>%
  filter(st_intersects(., 
                       shape_7901_buff, 
                       sparse = FALSE)) %>%
  arrange(study)
```

We'll define two segments at the queue jump signals that the 79 Southbound uses. Note that map layers can be toggled on and off in left panel and a measuring tool within the map can be used to check distances.

1. At [the piney branch intersection](https://www.google.com/maps/place/Piney+Branch+Rd+NW+%26+Georgia+Ave+NW,+Washington,+DC+20011,+USA/@38.9651607,-77.0298217,15.44z/data=!4m5!3m4!1s0x89b7c889f9d2cced:0x2f64ed3efceac785!8m2!3d38.9674749!4d-77.0274863), From Piney Branch in Yellow, it's about 395 feet to the previous intersection at Underwood, but the previous signalized intersection is much further back at Van Buren Street. All of the intersections shown between Piney Branch and next stop are signalized; though the intersection at Tuckerman St. is also signalized, the 79 does not 'hit' an additional light here in either the north or southbound direction. 

```{r message=FALSE, warning=FALSE}
map <- 
{mapview(intersections_7901,
        zcol = "study") + 
    mapview(stops_7901, col.regions = "red", layer.name = "79 SB stops") +
  mapview(shape_7901_buff, legend = FALSE) +
  mapview(shape_7901, legend = TRUE, layer.name = "79 SB")}@map %>%
  leaflet::addMeasure()

map$sizingPolicy$knitr$figure <- FALSE

map %>%
  leaflet::setView(lat = 38.96433, lng = -77.02779, zoom = 16)

```

2. At [Georgia Avenue and Columbia Road, SB](https://www.google.com/maps/place/Georgia+Ave+%2B+Columbia+Rd/@38.9284137,-77.0238868,18.16z/data=!4m13!1m7!3m6!1s0x89b7c86c33c6d3ed:0x387280dc1acad93!2sGeorgia+Ave+NW,+Washington,+DC,+USA!3b1!8m2!3d38.9486038!4d-77.0247807!3m4!1s0x89b7c81cd95eb033:0x9819589307a7732!8m2!3d38.9282781!4d-77.023228). Immediately prior to this queue jump is another queue jump at Irving Street. However, the stop at Irving Street is only used by Route 70.

```{r}
map %>%
  leaflet::setView(lat = 38.92828, lng = -77.02324, zoom = 16)
```

For Piney Branch, we'll just create a linestring based on coordinates picked off the map, roughly 300 ft before the QJ (just south of Underwood) and 1,500 feet after (half way between third and fourth intersection out). For Columbia Road, we'll create a segment starting north of Irving Street (another Queue Jump intersection) near Kenyon Street. 
```{r}
segments <- 
  tribble(~name, ~source,~lat, ~lon,
          "piney",  "start", 38.968452,-77.027389,
         "piney", "end" ,38.963196,-77.027872,
         "columbia_irving", "start", 38.92984, -77.02337,
         "columbia_irving", "end", 38.92331, -77.02242) %>%
  st_as_sf(., 
           coords = c("lon", "lat"),
           crs = 4326L, #WGS84
           agr = "constant") %>%
  group_by(name) %>%
  summarize() %>%
  st_cast("LINESTRING") %>%
  st_transform(DCCRS)

segments_buff <-
  segments %>%
  st_buffer(., dist = 50, endCapStyle = "FLAT")
```

Any case where points are continuously within this short buffer will be used for analysis. The Piney Branch and Columbia-Irving Segments are shown in turn.
```{r, out.width = '100%'}
map <- 
  {mapview(intersections_7901, layer.name = "Intersections") + 
    mapview(stops_7901, col.regions = "red", layer.name = "79 SB stops") +
  mapview(segments_buff, legend = TRUE, layer.name = "Analysis Segment")}@map %>%
  addMeasure() 

map$sizingPolicy$knitr$figure <- FALSE

map %>%
  leaflet::setView(lat = 38.96433, lng = -77.02779, zoom = 16)
```


```{r}
map %>%
  leaflet::setView(lat = 38.92677, lng = -77.02287, zoom = 16)

```


# Filter RawNav to data in these points

To save processing time, we'll filter to 79 southboundcases between several stops, then convert points to geometry, then filter to those in our study area. 

First, let's look once more at where the stop codes are since we don't have GTFS data joined in
```{r}
rawnav7901_preview <-
  rawnav79 %>%
  lazy_dt() %>%
  filter(route_pattern == "7901") %>%
  filter(filename == first(filename)) %>%
  filter(IndexTripStartInCleanData == first(IndexTripStartInCleanData)) %>%
  collect() %>%
  st_as_sf(., 
           coords = c("Long", "Lat"),
           crs = 4326L, #WGS84
           agr = "constant") %>%
  st_transform(DCCRS)
```

Looks like Piney Branch is stop E04, so we'll need points between stops E03 and E04 for Piney Branch. For Columbia Road, we'll need stop E09, so we'll grab E08 and E09 to be safe.
```{r message=FALSE, warning=FALSE}
map <-
  {
    mapview(segments_buff) +
      mapview(filter(rawnav7901_preview, str_detect(StopWindow, "E")),
              zcol = "StopWindow",
              legend = FALSE)
  }@map %>%
  leaflet::setView(lat = 38.96433,
                   lng = -77.02779,
                   zoom = 16)

map$sizingPolicy$knitr$figure <- FALSE

map
```

We'll create a stop segment field based on the stop a route has just *departed*, filter to points in our stop segments, convert to geometry, intersect against our buffer.

```{r}
rawnav79_segs_1 <-
  rawnav79 %>%
  lazy_dt() %>%
  filter(stop_segment %in% c("E03","E04","E08","E09")) %>%
  ungroup() %>%
  collect() %>%
  st_as_sf(., 
           coords = c("Long", "Lat"),
           crs = 4326L, #WGS84
           agr = "constant") %>%
  st_transform(DCCRS)

rawnav79_segs <-
  rawnav79_segs_1 %>%
  st_intersection(segments_buff)

```

Let's take a look at one trip through the Piney Branch segment.

```{r}
map <-
  {mapview(segments_buff, col.regions = "blue", legend = FALSE) +
      {rawnav79_segs %>%
          filter(name == "piney") %>%
          filter(IndexTripStartInCleanData == first(IndexTripStartInCleanData )) %>%
          filter(filename == first(filename)) %>%
          mapview(., zcol = "stop_segment")}}@map %>%
  leaflet::setView(lat = 38.96433,
                   lng = -77.02779,
                   zoom = 16)

map$sizingPolicy$knitr$figure <- FALSE

map
```

# Segment-level Speeds

To keep it simple, let's collapse now to the first and last point, calculate the odometer distance and time differences, and get some speeds.

```{r}
rawnav79_segs_run_sum_1 <-
  rawnav79_segs %>%
  st_drop_geometry() %>%
  group_by(name, filename,IndexTripStartInCleanData) %>%
  summarize(odo_seg_start = min(OdomtFt, na.rm = TRUE),
            odo_seg_end = max(OdomtFt, na.rm = TRUE),
            time_seg_start = min(SecPastSt, na.rm = TRUE),
            time_seg_end = max(SecPastSt, na.rm = TRUE),
            odo_seg_diff = odo_seg_end - odo_seg_start,
            time_seg_diff = time_seg_end - time_seg_start,
            ft_per_sec = odo_seg_diff / time_seg_diff,
            mph = as.numeric(ft_per_sec) / 1.467) #approximation because i'm a hack

```

We'll remove bus runs with potential data quality issues before calculating distributions. In particular, our Columbia-Irving and Piney Branch segments are `r paste(round(as.numeric(st_length(segments_buff)),2), collapse = " and")` feet long, respectively; we'll filter out any cases where the distance traveled was significantly longer or shorter than this.
```{r}
rawnav79_segs_run_sum_2 <-
  rawnav79_segs_run_sum_1 %>%
  filter((name == "piney" & (odo_seg_diff > (1919-200) & odo_seg_diff < (1919+200))) |
           (name == "columbia_irving" & (odo_seg_diff >(2394-200) & odo_seg_diff < (2394+200)))) %>%
  ungroup() %>%
  mutate(name = case_when(name == "columbia_irving" ~ "Columbia Rd",
                          name == "piney" ~ "Piney Branch Rd"))
```

The distribution of trip speeds is shown below.

```{r}
ggplot(rawnav79_segs_run_sum_2) +
  geom_histogram(aes(x = mph)) +
  facet_grid(rows = vars(name)) +  
  labs(x = "Average Segment Speed (mph)", y = "Count of Trips") + 
  FITP_Theme
```

Notably, there's a school slow zone somewhat south the Piney Branch intersection, and perhaps buses are slowing down later on, especially during school hours that might overlap with many of the trips in this dataset: [See google maps link here](https://www.google.com/maps/@38.9647409,-77.0277021,3a,41.1y,217.38h,85.85t/data=!3m6!1e1!3m4!1sPbn3DM4vSlkDSPt9ZM-mqA!2e0!7i16384!8i8192). Near Girard south of Columbia Road, there is also a number of crosswalks.

Using this dataset as a distribution, the range of free flow travel speeds is shown below. By comparison, the posted speed in these segments is 30 mph in the southbound direction.

```{r}
rawnav79_segs_run_sum_2 %>%
  group_by(name) %>%
  summarize(distr = list(quantile(mph, probs = c(.02,.05,.1,.25,.5,.75,.9,.95,.98)))) %>%
  unnest_wider(col = distr)
```

# Ping-to-Ping Speeds

As an alternative, lets look at the nth percentile of speed for each ping. This gets us closer to the posted speed limit in terms of speed distributions.

```{r}
rawnav79_segs_add_speed <-
  rawnav79_segs %>%
  group_by(name, filename,IndexTripStartInCleanData) %>%
  mutate(next_odo_feet_marginal = lead(OdomtFt, default = NA) - OdomtFt,
         next_seconds_marginal = lead(SecPastSt, default = NA) - SecPastSt,
         next_tot_seconds = lead(SecPastSt, default = NA),
         next_mph = (next_odo_feet_marginal / 5280) / (next_seconds_marginal / 3600))

ping_to_ping_distr <-
  rawnav79_segs_add_speed %>%
  filter(!is.infinite(next_mph)) %>%
  group_by(name) %>%
  summarize(next_mph_distr = list(quantile(next_mph, probs = c(.02,.05,.1,.25,.5,.75,.9,.95,.98), na.rm = TRUE))) %>%
  unnest_wider(col = next_mph_distr) %>%
  select(-geometry)

ping_to_ping_distr
```

