---
title: "TStop2 cut Content"
output: html_notebook
---


Rather than trying to move back the segment start for now, we'll use the existing start of the segment and go past the same amount (about 400 feet); in this way, we'll have a sort of V-shaped interval to look at.

```{r}
#note, accel/decel stuff abbreviated ad
segment_piney_ad <- 
  tribble(~source,~lat, ~lon,
         "start", 38.968452,-77.027389,
         "end" ,38.966541,-77.027580) %>%
  st_as_sf(., 
           coords = c("lon", "lat"),
           crs = 4326L, #WGS84
           agr = "constant") %>%
  st_transform(DCCRS)
```

First, we'll find the first and last points for each run on this shorter segment. a sample is shown below
```{r}

rawnav79_piney_marg_ad <-
  rawnav79_piney_marg %>%
  ungroup()%>%
  #we didn't calculate this for the entire sample, so we have NAs for the first 5
  #of each run within the segment
  filter(!is.na(next_mph_window)) %>%
  mutate(dist_to_start_ad = st_distance(geometry,
                                        filter(segment_piney_ad,source == "start")),
         dist_to_end_ad = st_distance(geometry,
                                      filter(segment_piney_ad,source == "end"))) %>%
  group_by(filename,IndexTripStartInCleanData) %>%
  mutate(ad_start = dist_to_start_ad == min(dist_to_start_ad),
         ad_end  = dist_to_end_ad == min(dist_to_end_ad),
         ad_class = case_when(ad_start ~ "start",
                              ad_end ~ "end")) %>%
  filter(!is.na(ad_class)) %>%
  #could have ties if rawnav not moving, so we take the first of each case
  group_by(filename,IndexTripStartInCleanData,ad_class) %>%
  filter(row_number() ==1) %>%
  ungroup()

#check start and end for each
rawnav79_piney_marg_ad %>% 
  count(filename,IndexTripStartInCleanData) %>% 
  pull(n) %>%
  {all(. == 2)} %>%
  testthat::expect_true()

rawnav79_piney_marg_ad %>% st_drop_geometry() %>% head(n=20)
```

Next we'll summarize these to get starting speed, expected runtime, actual runtime, and the difference -- Tstop2
```{r}
rawnav79_piney_marg_ad_sum <-
  rawnav79_piney_marg_ad %>%
  st_drop_geometry() %>%
  select(filename, IndexTripStartInCleanData,OdomtFt, SecPastSt,next_mph,next_mph_window, ad_class) %>%
  pivot_wider(names_from = ad_class, values_from = (OdomtFt:next_mph_window)) %>%
  mutate(tt_actual = SecPastSt_end - SecPastSt_start,
         odom_actual = OdomtFt_end - OdomtFt_start,
         fps_actual = odom_actual / tt_actual,
         mph_actual = (odom_actual / 5280) / (tt_actual / 3600),
         #convert from next_mph_window_start
         fps_nodelay = next_mph_window_start * (5280 / 3600),
         tt_nodelay = odom_actual / fps_nodelay,
         tstop2 = tt_actual - tt_nodelay)
```

So what's this get us? Some values are negative -- if we happen to get a slow speed at the start of the intersection, it's possible the bus later speeds up. 

```{r}
rawnav79_piney_marg_ad_sum %>%
  ggplot(aes(x = tstop2)) + 
  geom_histogram()
```

```{r}
rawnav79_piney_marg_ad_sum$tstop2 %>% summary()
```

Let's look back at that chart, filtering only to cases with negative tstop2 delay.

Indeed, it looks like these are cases that are nearly at a stop at the start of the segment, accelerate, and then decelerate to hit the stop before the QJ, then accelerate again.

```{r, out.width='100%'}
funcases <-
  rawnav79_piney_marg_ad_sum %>%
  filter(tstop2 <= 0) %>%
  pull(busrun)


plotdf <-
  rawnav79_piney_marg %>%
  filter(busrun %in% funcases) %>%
  plotly::highlight_key(., ~busrun)
    
plot <-
  ggplot(plotdf,
         aes(x = OdomtFt_seg, y = next_mph_window, group = busrun)) + 
  geom_line()

gg <- ggplotly(plot, tooltip = "busrun")

highlight( gg, on = "plotly_hover", off = "plotly_deselect", color = "red" )

```